A Short Guide to Anyons and Modular Functors

### 2016-10-17 guide.tex
more tweaking

In the physics literature, modular functors are
explicitly used in Refs.~\cite{Freedman2002, Freedman2002simulation}.
Also, \cite{Beverland2014} and \cite{Kitaev2006topo}
use the language of modular functors but they call them TQFT's.
This is in fact reasonable because a modular functor can be
seen as part of a TQFT, but is quite confusing to the
novice who attempts to delve into the mathematical literature.

@@

What we have given is a topological description of the group $B_n$.
More formally, we can describe these braid world lines as
paths in the \emph{configuration space} of $n$ points.
\newcommand{\Conf}{\mathcal{C}}
\newcommand{\UConf}{\mathcal{UC}}
This space is defined as the product of
a two dimensional space $M$ for each point, minus the subspace where points overlap:
$$
    \Conf_n = \bigl( \prod_{1}^{n} M \bigr) - \Delta, \ \ 
    \Delta = \{(x_1, ..., x_n) | x_i = x_j \ \ \mbox{for some} \ \ i\ne j \}.
$$
Because we are considering identical particles (so far)
we use the \emph{unlabelled configuration space}
$\UConf_n$ which is the quotient of $\Conf_n$ by the natural action of the
permutation group $S_n:$
%\cggb{I was a bit confused by when we care about
%unlabelled configuration space and when we care about the labelled
%configuration space. What if we have some but not all
%particles identical? Do the two formalisms unify by considering some
%subgroups of the permutation group that preserve particle label or something?}
$$
    \UConf_n = \Conf_n / S_n.
$$
The \emph{geometric braid group} as we have so far informally described it 
can now be rigorously defined as the fundamental group:
$$
    B_n = \pi_1 ( \UConf_n, x )
$$
where $x$ is some reference configuration in $\UConf_n.$
See \cite{Ghrist2014} for further discussion.

@@

There is another important geometric representation of
the braid group which is purely two-dimensional.
We pick an $n$-point finite subset $Q_n \subset M$
and consider diffeomorphisms 
%\cggb{why diffeomorphisms and not homeomorphisms?}
$f : M \to M$
that map $Q_n$ to itself.
The \emph{mapping class group} of $M$ relative to $Q_n$
is the set of all such diffeomorphisms up to an equivalence relation $\sim_{\mbox{iso}}$:
$$
    MCG(M, Q_n) = \{ f : M \to M \ \ \mbox{such that}\ \ f(Q_n)=Q_n \} / \sim_{\mbox{iso}}
$$
The equivalence relation $\sim_{\mbox{iso}}$ is called \emph{isotopy}
which allows for any continuous deformation of $f:M\to M$ that
fixes each point in $Q_n.$
Each generator $\sigma_i$ of the braid group is found in $MCG(M, Q_n)$
as a \emph{half-twist} that swaps two points $i, i+1 \in Q_n$:
\begin{center}
\includegraphics[]{pic-halftwist.pdf}
\end{center}
In order to show
the action of
this half-twist we have 
decorated the manifold with
a a checkered pattern,
but there is a more important
object that lives on the manifold
itself.
An \emph{observable} is a simple
closed curve in $M$ that does
not intersect $Q_n.$
Such a closed curve is called an
observable because these will be
associated to measurements
that measure the charge on the
interior of the curve.
The importance of understanding the
braid group as identical to the mapping class group
is now manifest:
whereas geometric braids act on states
as in a Schr\"{o}dinger picture,
elements of the mapping class group
act on the observables as in
a Heisenberg picture.
\begin{center}
\includegraphics[]{pic-heisenberg.pdf}
\end{center}
This diagram should give the reader some idea as
to why these two definitions of the braid group are
equivalent, but the actual proof of this is somewhat involved.
We refer to \cite{Kassel2010} for an excellent contemporary
reference that fills in these gaps.

@@

%By an \emph{isotopy} of $f:M\to N$ we mean a 
%map $g:M\to N$
%that is a continuos deformation of $f.$
Two maps of surfaces $f:M\to N$ and $f':M\to N$
will be called \emph{isotopic}
when one is a continuous deformation of the other.
In detail,
we have a continuously
parametrized family of maps $f_t:M\to N$ for $t\in [0, 1]$
such that $f_0=f,\ f_1=f'$
and the restriction of $f_t$ to the 
set of marked points $X\subset\partial M$ is constant:
$f_t|_{X} = f|_{X}$ for $t\in [0, 1].$
%and the restriction of $f_t$ to the boundary is constant:
%$f_t|_{\partial M} = f|_{\partial M}$ for $t\in [0, 1].$
Such a family $\{f_t\}_{t\in [0,1]}$ is called an \emph{isotopy}
of $f.$
This is an equivalence relation on maps $M\to N$, and the equivalence
class of $f$ under isotopy is called the \emph{isotopy class} of $f.$
%\cggb{I didn't quite see why this more formal definition is
%the same as equivalent up to continuous deformation. Why can
%I restrict to maps where the base points are fixed along the whole path?}

@@

\subsub{POP decomposition.}
%The reverse process of cutting a surface up
%into pair-of-pants induces a direct sum decomposition
%of the fusion space.
Given a manifold $M$ which is
a disc with two or more holes, 
we show how to present $M$ as the gluing of various
pair-of-pants. Such an arrangement will be termed a \emph{POP decomposition.}
(We refer to \cite{Ivanov2001} for more details on this construction,
and \cite{Ghrist2014} for a leisurely description of Morse theory.)
This will yield a decomposition of $\F(M)$ into a direct
sum of fusion spaces of pair-of-pants. % $V_{ab}^c$'s.
%The pair-of-pants can be found using Morse theory.
%We proceed using Morse theory.
%Note that we need $M$ to be a differentiable manifold for
%this to work.
The key idea is to choose a ``height'' function $h:M\to\R$
with some specific properties that allow us to
cut the manifold up along level sets of $h.$
First, we need that critical points of $h$ are isolated.
This is the defining condition for $h$ to be a \emph{Morse function}.
%Choose a Morse function (a generic ``height'' function) $h:M\to\R$
Also, we need that $h$ is constant on $\partial M,$
and the values of $h$ at different critical points are distinct.
Now choose a sequence of non-critical values $a_1<a_2<...<a_n$ in $\R$
such that every interval $[a_{i-1}, a_i]$ 
contains exactly one critical value of $h$ and the image
of $h$ lies within $[a_1, a_n].$
%cuts the manifold into pieces $f^{-1}[a,b]$.
Each component of $h^{-1}([a_{i-1}, a_i])$ is then either
an annulus, a disc, or a pair-of-pants depending on the index of (any)
critical point it contains.
We then re-glue any annuli or discs until there are no more of these
and we have only pair-of-pants.
%\cggb{Is this always possible? What if I have something like
%a cylinder? Then I just get a bunch of annuli however I cut.}
% note: we may still get annuli in a *component* of the pre-image.
\begin{center}
\includegraphics[]{pic-pants.pdf}\ \ \ \ \ \ \ \ \includegraphics[]{pic-pants-1.pdf}
\end{center}

@@

One way to achieve this is via Cerf theory, which is the theory
of how one may deform Morse functions into other Morse functions
and the kind of transitions involved in their critical point structure.
This was the approach used in \cite{Freedman2002simulation}.
In this work we use a simpler method,
which is essentially the same as {skein theory.}
%    \cggb{Is all this curve diagrams and refactoring theorem and and
%    so on original work, or is it collecting known facts?
%    If its original, this should be an independent paper because
%    otherwise nobody from the math phys community will ever find
%    it. Actually maybe in either case it should be an
%    independent arxiv post at least so that more people find and learn from it.}
%Read on to find out how this works!
This is the {refactoring theorem} that we describe below.

@@

Each standard surface comes with a collection of
\emph{standard POP decompositions}:
these will be POP decompositions where we require each
observable to cross the
equator twice and have counterclockwise orientation.
Up to isotopy, a given standard surface will have only finitely many of these.
On a standard surface with three
interior holes there are two standard POP decompositions,
and one $F$-move that relates these.
On a standard surface with four interior holes there are five 
standard POP decompositions and five $F$-moves that relate these.
In this case the $F$-moves themselves satisfy an equation
that is a trivial consequence of the way we have defined $F$-moves.
%enforced by the axioms of the modular functor.
This is known as the pentagon equation, which we depict as
the following commutative diagram:
\begin{center}
\includegraphics[]{pic-pentagon.pdf}
\end{center}
We are now starting to confuse the notation 
for the topological space $M$ and the fusion space $\F(M).$
Also, each of these $F$-moves is referring to an isomorphism 
that is block decomposed according to the charge sectors 
of the indicated observables.

@@

@@

>>>>

In the physics literature, modular functors are
explicitly used in Refs.~\cite{Freedman2002, Freedman2002simulation}.
Also, Refs.~\cite{Beverland2014} and \cite{Kitaev2006topo}
use the language of modular functors but they call them TQFT's.
This is in fact reasonable because a modular functor can be
seen as part of a TQFT, but is quite confusing to the
novice who attempts to delve into the mathematical literature.

@@

What we have given is a topological description of the group $B_n$.
More formally, we can describe these braid world lines as
paths in the \emph{configuration space} of $n$ points.
\newcommand{\Conf}{\mathcal{C}}
\newcommand{\UConf}{\mathcal{UC}}
This space is defined as the product of
a two dimensional space $M$ for each point, minus the subspace where points overlap:
$$
    \Conf_n = \bigl( \prod_{1}^{n} M \bigr) - \Delta, \ \ 
    \Delta = \{(x_1, ..., x_n) | x_i = x_j \ \ \mbox{for some} \ \ i\ne j \}.
$$
Because we are considering identical particles (so far)
we use the \emph{unlabelled configuration space}
$\UConf_n$ which is the quotient of $\Conf_n$ by the natural action of the
permutation group $S_n:$
%\cggb{I was a bit confused by when we care about
%unlabelled configuration space and when we care about the labelled
%configuration space. What if we have some but not all
%particles identical? Do the two formalisms unify by considering some
%subgroups of the permutation group that preserve particle label or something?}
$$
    \UConf_n = \Conf_n / S_n.
$$
The \emph{geometric braid group} as we have so far informally described it 
can now be rigorously defined as the fundamental group:
$$
    B_n = \pi_1 ( \UConf_n, x )
$$
where $x$ is some reference configuration in $\UConf_n.$
See Ref.~\cite{Ghrist2014} for further discussion.

@@

There is another important geometric representation of
the braid group which is purely two-dimensional.
We pick an $n$-point finite subset $Q_n \subset M$
and consider diffeomorphisms 
%\cggb{why diffeomorphisms and not homeomorphisms?}
$f : M \to M$
that map $Q_n$ to itself.
The \emph{mapping class group} of $M$ relative to $Q_n$
is the set of all such diffeomorphisms up to an equivalence relation $\sim_{\mbox{iso}}$:
$$
    MCG(M, Q_n) = \{ f : M \to M \ \ \mbox{such that}\ \ f(Q_n)=Q_n \} / \sim_{\mbox{iso}}
$$
The equivalence relation $\sim_{\mbox{iso}}$ is called \emph{isotopy}
which allows for any continuous deformation of $f:M\to M$ that
fixes each point in $Q_n.$
Each generator $\sigma_i$ of the braid group is found in $MCG(M, Q_n)$
as a \emph{half-twist} that swaps two points $i, i+1 \in Q_n$:
\begin{center}
\includegraphics[width=1.0\columnwidth]{pic-halftwist.pdf}
\end{center}
In order to show
the action of
this half-twist we have 
decorated the manifold with
a a checkered pattern,
but there is a more important
object that lives on the manifold
itself.
An \emph{observable} is a simple
closed curve in $M$ that does
not intersect $Q_n.$
Such a closed curve is called an
observable because these will be
associated to measurements
of the total anyonic charge on the
interior of the curve.
The importance of understanding the
braid group as identical to the mapping class group
is now manifest:
whereas geometric braids act on states
as in a Schr\"{o}dinger picture,
elements of the mapping class group
act on the observables as in
a Heisenberg picture.
\begin{center}
\includegraphics[]{pic-heisenberg.pdf}
\end{center}
This diagram should give the reader some idea as
to why these two definitions of the braid group are
equivalent, but the actual proof of this is somewhat involved.
We cite Ref.~\cite{Kassel2010} for an excellent contemporary
account that fills in these gaps.

@@

%By an \emph{isotopy} of $f:M\to N$ we mean a 
%map $g:M\to N$
%that is a continuos deformation of $f.$
Two maps of surfaces $f:M\to N$ and $f':M\to N$
will be called \emph{isotopic}
when one is a continuous deformation of the other.
In detail,
we have a continuously
parametrized family of maps $f_t:M\to N$ for $t\in [0, 1]$
such that $f_0=f,\ f_1=f'$
and the restriction of $f_t$ to the 
set of marked points $X\subset\partial M$ is constant:
$f_t|_{X} = f|_{X}$ for $t\in [0, 1].$
%and the restriction of $f_t$ to the boundary is constant:
%$f_t|_{\partial M} = f|_{\partial M}$ for $t\in [0, 1].$
Such a family $\{f_t\}_{t\in [0,1]}$ is called an \emph{isotopy}
of $f.$
This is an equivalence relation on maps $M\to N$, and the equivalence
class of $f$ under isotopy is called the \emph{isotopy class} of $f.$
%\cggb{I didn't quite see why this more formal definition is
%the same as equivalent up to continuous deformation. Why can
%I restrict to maps where the base points are fixed along the whole path?}
The weaker notion of \emph{homotopy} of maps will not be used
here, but for the maps we use
it turns out that homotopy is equivalent to isotopy.
Furthermore, we can weaken the requirement that maps be differentiable,
because every continuous map $f:M\to N$ is (continuously)
isotopic to a differentiable map~\cite{Farb2011}.

@@

\subsub{POP decomposition.}
%The reverse process of cutting a surface up
%into pair-of-pants induces a direct sum decomposition
%of the fusion space.
Given a manifold $M$ which is
a disc with two or more holes, 
we show how to present $M$ as the gluing of various
pair-of-pants. Such an arrangement will be termed a \emph{POP decomposition.}
(We refer to Ref.~\cite{Ivanov2001} for more details on this construction,
and Ref.~\cite{Ghrist2014} for a leisurely description of Morse theory.)
This will yield a decomposition of $\F(M)$ into a direct
sum of fusion spaces of pair-of-pants. % $V_{ab}^c$'s.
%The pair-of-pants can be found using Morse theory.
%We proceed using Morse theory.
%Note that we need $M$ to be a differentiable manifold for
%this to work.
The key idea is to choose a ``height'' function $h:M\to\R$
with some specific properties that allow us to
cut the manifold up along level sets of $h.$
First, we need that critical points of $h$ are isolated.
This is the defining condition for $h$ to be a \emph{Morse function}.
%Choose a Morse function (a generic ``height'' function) $h:M\to\R$
Also, we need that $h$ is constant on $\partial M,$
and the values of $h$ at different critical points are distinct.
Now choose a sequence of non-critical values $a_1<a_2<...<a_n$ in $\R$
such that every interval $[a_{i-1}, a_i]$ 
contains exactly one critical value of $h$ and the image
of $h$ lies within $[a_1, a_n].$
%cuts the manifold into pieces $f^{-1}[a,b]$.
Each component of $h^{-1}([a_{i-1}, a_i])$ is then either
an annulus, a disc, or a pair-of-pants depending on the index of (any)
critical point it contains.
We then re-glue any annuli or discs until there are no more of these
and we have only pair-of-pants.
%\cggb{Is this always possible? What if I have something like
%a cylinder? Then I just get a bunch of annuli however I cut.}
% note: we may still get annuli in a *component* of the pre-image.
\begin{center}
\includegraphics[]{pic-pants.pdf}\ \ \ \ \ \ \ \ \includegraphics[]{pic-pants-1.pdf}
\end{center}

@@

One way to achieve this is via Cerf theory, which is the theory
of how one may deform Morse functions into other Morse functions
and the kind of transitions involved in their critical point structure.
This was the approach used in Ref.~\cite{Freedman2002simulation}.
In this work we use a simpler method,
which is essentially the same as {skein theory.}
%    \cggb{Is all this curve diagrams and refactoring theorem and and
%    so on original work, or is it collecting known facts?
%    If its original, this should be an independent paper because
%    otherwise nobody from the math phys community will ever find
%    it. Actually maybe in either case it should be an
%    independent arxiv post at least so that more people find and learn from it.}
%Read on to find out how this works!
This is the {refactoring theorem} that we describe below.

@@

Each standard surface comes with a collection of
\emph{standard POP decompositions}:
these will be POP decompositions where we require each
observable to cross the
equator twice and have counterclockwise orientation.
Up to isotopy, a given standard surface will have only finitely many of these.
On a standard surface with three
interior holes there are two standard POP decompositions,
and one $F$-move that relates these.
On a standard surface with four interior holes there are five 
standard POP decompositions and five $F$-moves that relate these.
In this case the $F$-moves themselves satisfy an equation
that is an immediate consequence of the way we have defined $F$-moves.
%enforced by the axioms of the modular functor.
This is known as the pentagon equation, which we depict as
the following commutative diagram:
\begin{center}
\includegraphics[]{pic-pentagon.pdf}
\end{center}
We are now starting to confuse the notation 
for the topological space $M$ and the fusion space $\F(M).$
Also, each of these $F$-moves is referring to an isomorphism 
that is block decomposed according to the charge sectors 
of the indicated observables.

@@

% See also: Lantern relation \cite{Farb2011},
% involving seven Dehn twists on the sphere with 4 holes.

@@

\begin{acknowledgments}

The author wishes to thank 
Andrew Doherty, Courtney Brell, Steven Flammia and Dominic Williamson
for helpful discussions and inspiration.

\end{acknowledgments}
